CMSC818s: Project 3 \\ A Brief Survey of Modern Virtualization Research

### 2011-12-13 paper.tex
small edits

Virtualization got it's start back in the 1960s, when developers
of large time-sharing systems sought ways to give each user the illusion that
they had sole access to a private, virtual computer, despite the fact that
many users were in fact sharing computing resources.  Once private physical
computers became available, popular, and cheap in the 80s, virtualization
took a backseat to other areas of research.  However, many of the ideas
put forth in these early days lived on in work involving virtual memory
and operating system protection paradigms.

Server virtualization arrived on the scene about a decade ago, with VMWare
appearing in 1999 and Xen in 2003.  Since then, research and commercial interest
in virtualization has sparked significantly, especially in conjunction
with the rise of Cloud computing and Infrastructure-as-a-service.

In a virtualized context, we have many \emph{virtual machines}(VM) sharing the same
physical hardware.  In a \emph{fully virtualized} system, each VM has the illusion
that it alone has sole access to the hardware, although in a \emph{paravirtualized}
system VMs may be modified to be aware of the fact that they are virtualized.  This
sharing of resources is possible because access to the hardware is regulated by
a special piece of software, the \emph{virtual machine monitor} or \emph{hypervisor}.
In many ways, the hypervisor plays the role of the kernel in a traditional operating
system - it contains the privileged software that can directly access the hardware.

@@

Three years after VMWare's appearance, Denali \cite{ref:denali} sought to provide a virtualization
solution design with two goals in mind, scalability and security. Denali's authors
envisioned a separation between the management of physical hardware and the management
of software services, transforming how internet services could be deployed. Such a system 
would allow untrusted software to run on third party infrastructure, support the distribution 
of dynamic content in delivery systems, or provide virtual hosting infrastructure.

In order to meet this vision, the Denali project introduced \emph{paravirtualization} by providing
a simplified and pared-down emulation of the hardware for VMs to interface with.  Denali also
adopted complete isolation of guest VMs, and made VMs extremely lightweight, such that the system can
support over 10,000 VMs on commodity machines. Denali diverges from many 
related projects by sacrificing backwards compatibility for its primary objectives,
but its designers argue that this design tradeoff is well worth the benefits.

@@

Denali's virtual architecture and implementation adhere to a separation between mechanism and policy 
so changes can be easily made. CPU virtualization holds a policy split into two elements, 
a \emph{gatekeeper policy} to choose the subset of active VMs to admit into the system, and 
a \emph{scheduler policy} to switch between active VMs. For simplicity, the authors use FIFO
and round robin in the descirbed implementation, although other scheduling policies could
be implemented.  The system provides a non-traditional interrupt model, where virtual interrupts
for a VM are batched and delivered asynchronously when the VM is next scheduled.

@@

Denali also provides some limited emulated hardware.  In particular, a virtual ethernet NIC is 
provided for each VM, with queues for incoming and outgoing packets.
Incoming packets for each VM are accumulated in the appropriate queues and are later processed once the
corresponding VM is given focus by the scheduler.  Outgoing packets are processed in a
round-robin manner from outgoing queues. Denali's TCP/IP stack runs at the user level, which
has its drawbacks, since all packets have to cross the user/kernel boundary. However, the paper contains
a packet dispatch evaluation, which suggests that cost of network virtualization is low and that the the
driver itself is responsible for the majority of the costs of packet processing.

@@

Key contributions of this work include the design and implementation
of a truly efficient nested VM for the x86 architecture, including novel
techniques for \emph{multidimensional paging} and
\emph{multi-level device assignment}.  The Turtles hypervisor is able to
run unmodified guest hypervisors and OSes with a performance penalty of
as low as 6-8\%.  The paper also presents what the authors claim is the
first thorough analysis and evaluation of nested hypervisor performance,
with an eye to the identification and reduction of performance bottlenecks.

@@

\begin{figure}[t]
	\begin{center}
		\includegraphics[width=2.5in]{images/mclock.png}
	\end{center}
	\caption{The mClock algorithm works by tagging requests with a tag for
		reservations, weights, and reservations, and then executing requests
		in order of their tags.  (Figure courtest of \cite{ref:mclock})}
	\label{fig:mclock-alg}
\end{figure}

@@

In actuality, things are a little bit more complicated.  mClock maintains
3 types of tags and 3 global clocks, one each for reservations, weights,
and limits.  Each iteration of the algorithm is deemed to be
either a ``constraint-based phase'', in which some VM has not received
capacity above its reservation, or a ``weight-based phase'', in which the
next VM to be scheduled will be chosen by proportional share.  This is easily
determined after the tagging:  if any VM $i$ has request $j$ such that the
reservation tag $R(i, j)$ is less than the current time, than VM $i$ is currently
receiving less capacity than its reservation, and so the current phase is
constraint based.  Otherwise, the phase is weight-based, and so the algorithm
just picks a request with minimum weight tag $P(i, j)$ whose limit tag
$L(i, j)$ is also less than the current time (otherwise, VM $i$ is receiving
more capacity than its limit, and so it should not be scheduled).

@@

Our papers are clearly divided by this metric:  Turtles is all about efficiency
and performance, often at the cost of fairness.  The problem seems
especially hard here when phrased in the nested context.  It seems perfectly logical that if
a guest OS running in $L_2$ should only receive a fraction $\alpha$ of the resources allocated
to its host in $L_1$, which in turn should only receive a percentage $\beta$ of the machine's
total resources, then these two fractions should compound and our guest should end up with
$\alpha\beta$ of the total capacity available.  However, in reality Turtles multiplexes
all guests at all levels onto $L_0$, so in many ways the guest OS and its hypervisor will
be treated equally, rather than as parent and child.  Multi-level device assignment
for Port I/O and Memory-mapped I/O is particularly agnostic to fairness concerns, because
it works exactly the same way as single-level device assignment, and so doesn't care whether
a guest runs at $L_1$ or $L_{100}$ - all guests get equal treatment.  Furthermore, no
experimentation at all is done to analyze the effects of this model on fairness;
the authors even make the statement that this is a secondary concern,
and so they are content with a sort of best-effort.

@@

\subsection{Paravirtualization and Full Virtualization}
\label{sec:relns/para-full}

The second tension seems to arise between the paradigms of \emph{paravirtualization}
and \emph{full virtualization}, which were introduced in Section \ref{sec:background}.
Corporately-funded literature seems to have an eye more toward using as much
commodity software as possible here, for the sake of lower adoption costs, while
academic papers seem more willing to consider modifying client software in
appropriate situations.

The split here is pretty clear as well: Denali, Xen, and Time all consider
paravirtualization to be beneficial and even necessary in the right situations.
In Denali, the isolation kernel presents a modified and simplified hardware abstraction
to its clients, which cleverly allows for some big wins in efficiency and performance
as a side effect.  Xen claims that paravirtualization is necessary to ensure fairness
and prevent starvation in process and I/O scheduling; one could argue, however, that
this is no longer a concern due to the advent of newer virtualization-aware hardware
(see Section \ref{sec:relns/hard-emul}).  The Time paper goes a step further and claims that
even with the new hardware, making the guest OS aware of the virtualized context it
runs in is necessary to ensure the correctness of certain time-dependent applications,
such as systems measurement, high-speed trading, and even video gaming.

In contrast, the Turtles project argues for full virtualization and unmodified guest
OSes and hypervisors.  In the paper, the argument for this goes back to the
motivating goal of performance, though it was unclear to us why paravirtualization
would have any negative impact here.  But it seems also that for any sort of large-scale
commercial deployment of server virtualization technology, the ability to use commodity
guests would be crucial in order to make adoption easy, and we consider this to also
be a likely factor influencing the design choice.

The mClock paper doesn't really fit into these two classifications, since it mostly
deals with the interactions between the hypervisor and its devices, rather than the hypervisor
and its guests.  However, the paper does mention that one of the reasons this disk
scheduling problem is so difficult is because guest OSes have their own algorithms
for efficient disk scheduling, which generally don't play nice with each other.  If it
were possible to modify the guest OS to disable these algorithms, so that the hypervisor
could handle all disk scheduling decision, the difficulty of this problem could likely
be reduced.  This seems to be an argument supportive of paravirtualization.

@@

The final tension we wish to consider is the question of scalability and designing
for the Cloud.  We have observed
that there is a commercial interest in developing super-scalable server virtualization,
largely because it has found such a unique niche with the market growth of Cloud computing
and Infrastructure-as-a-service (IAAS).  While academic research is not uninterested in the question
of scale or the Cloud use case, we do not feel that they are as much of a driving factor in this realm.

We see this trend in the papers we have been discussing as well.  The Turtles paper presents
the motivation for nested virtualization using the argument of its suitability for
Infrastructure-as-a-service, as the two are essentially made for each other.  The mClock paper,
although it does not particularly market itself
for scalability, is motivated by the Cloud-enabled scenario of many virtualized clients
needing to schedule the use of a shared network disk.  Furthermore, the guarantees mClock
makes about fairness, reservations, and proportional share are especially important
in this scenario, as they allow IAAS providers to keep service-level agreements with
their clients.  We find this particularly interesting, as it seems to indicate
that in the Cloud, fairness and performance may find themselves on equal footing in
terms of importance, despite what we have observed in Section \ref{sec:relns/perf-fair}.

Interestingly, the Denali paper also has a particular eye to scalability, and a large
portion of the evaluations section is devoted to demonstrating that the isolation kernel
performs well even when supporting over 10,000 VMs.  However, even in Denali, scalability
is just one goal amongst many, and the paper does not really consider applications
to Cloud computing, being written several years before the Cloud really became prominent.

@@

Having examined the recent literature along these four axes of tension, it is our belief
that a clear trend has emerged in virtualization research.  We believe that as Cloud computing and
IAAS continue to take off, research in this area will continue to trend toward what
is both popular and financially successful.  We expect that the majority of virtualization
research will continue in the direction of full virtualization over paravirtualization,
and that the Cloud-computing/IAAS scenario will become the driving use case.
However, we think that as a result, fairness and performance will no longer be
competing ends, but will in fact become equal goals in virtualized system design.
This will be necessary in order for Cloud providers to meet their SLAs to their
consumers, and we believe fair-scheduling and resource-isolation research will move
from the shadows to the forefront of both corporate and academic research.

>>>>

Virtualization techniques arose more than half a century ago in the early 1960s, when developers
of large time-sharing systems sought ways to give each user the illusion that
they had sole access to a private, virtual computer, despite the fact that
many users were in fact sharing computing resources.  Once private physical
computers became available, popular, and cheap in the 80s, virtualization
took a backseat to other areas of research.  However, many of the ideas
put forth in these early days lived on in work involving virtual memory
and operating system protection paradigms.

Server virtualization arrived on the scene about a decade ago, with VMWare
appearing in 1999, Denali in 2002, and Xen in 2003.  Since then, research and commercial interest
in virtualization has sparked significantly, especially in conjunction
with the rise of Cloud computing and Infrastructure-as-a-service.

In a virtualized context, we have many \emph{virtual machines}(VM) sharing the same
physical hardware.  In a \emph{fully virtualized} system, each VM has the illusion
that it alone has sole access to the hardware, although in a \emph{para-virtualized}
system VMs may be modified to be aware of the fact that they are virtualized.  This
sharing of resources is possible because access to the hardware is regulated by
a special piece of software, the \emph{virtual machine monitor} or \emph{hypervisor}.
In many ways, the hypervisor plays the role of the kernel in a traditional operating
system - it contains the privileged software that can directly access the hardware.

@@

Three years after VMWare's appearance, Denali \cite{ref:denali} sought to provide a virtualization
solution with two goals in mind, scalability and security. Denali's authors
envisioned a separation between the management of physical hardware and the management
of software services, transforming how internet services could be deployed. Such a system 
would allow untrusted software to run on third party infrastructure, support the distribution 
of dynamic content in delivery systems, or provide a foundation for virtual hosting.

In order to meet this vision, the Denali project introduced \emph{para-virtualization} by providing
a simplified and pared-down emulation of the hardware for VMs to interface with.  Denali also
adopted complete isolation of guest VMs, and made VMs extremely lightweight, so that the system could
support over 10,000 VMs on commodity machines. Denali diverged from many 
related projects by sacrificing backwards compatibility for its primary objectives,
but its designers argued that this design tradeoff was well worth the benefits.

@@

Denali's virtual architecture and implementation adhere to a separation between mechanism and policy 
so changes can be easily made. CPU virtualization holds a policy split into two elements, 
a \emph{gatekeeper policy} to choose the subset of active VMs to admit into the system, and 
a \emph{scheduler policy} to switch between active VMs. For simplicity, the authors use FIFO
and round robin in the described implementation, although other scheduling policies could
be implemented.  The system provides a non-traditional interrupt model, where virtual interrupts
for a VM are batched and delivered asynchronously when the VM is next scheduled.

@@

Denali also provides some limited emulated hardware.  In particular, a virtual ethernet NIC is 
provided for each VM, with queues for incoming and outgoing packets.
Incoming packets for each VM are accumulated in the appropriate queues and are later processed once the
corresponding VM is given focus by the scheduler.  Outgoing packets are processed in a
round-robin manner from outgoing queues. Denali's TCP/IP stack runs at the user level, which
has its drawbacks, since all packets have to cross the user/kernel boundary. However, the paper contains
a packet dispatch evaluation, which suggests that cost of network virtualization is low and that the the
driver itself is responsible for the majority of packet processing costs.

@@

Key contributions of this work include the design and implementation
of a truly efficient nested VM for the x86 architecture, including novel
techniques for \emph{multidimensional paging} and
\emph{multi-level device assignment}.  The Turtles hypervisor is able to
run unmodified guest hypervisors and OSes with a performance penalty
as low as 6-8\%.  The paper also presents what the authors claim is the
first thorough analysis and evaluation of nested hypervisor performance,
with an eye to the identification and reduction of performance bottlenecks. % not sure about ``with an eye to''

@@

\begin{figure}[ht!]
	%\begin{center}
		\includegraphics[width=2.65in]{images/mclock.png}
	%\end{center}
	\caption{The mClock algorithm works by tagging requests with a tag for
		reservations, weights, and reservations, and then executing requests
		in order of their tags.  (Figure courtest of \cite{ref:mclock})}
	\label{fig:mclock-alg}
\end{figure}

@@

In actuality, things are a little bit more complicated.  mClock maintains
3 types of tags and 3 global clocks, one each for reservations, weights,
and limits.  Each iteration of the algorithm is deemed to be
either a ``constraint-based phase'', in which some VM has not received
capacity above its reservation, or a ``weight-based phase'', in which the
next VM to be scheduled will be chosen by proportional share.  This is easily
determined after the tagging:  if any VM $i$ has request $j$ such that the
reservation tag $R(i, j)$ is less than the current time, then VM $i$ is currently
receiving less capacity than its reservation, and so the current phase is
constraint based.  Otherwise, the phase is weight-based, and so the algorithm
just picks a request with minimum weight tag $P(i, j)$ whose limit tag
$L(i, j)$ is also less than the current time (otherwise, VM $i$ is receiving
more capacity than its limit, and so it should not be scheduled).

@@

Our papers are clearly divided by this metric:  Turtles is all about efficiency
and performance, often at the cost of fairness.  The problem seems
especially hard here when phrased in the nested context.  It seems perfectly logical that
a guest OS running in $L_2$ should only receive a fraction $\alpha$ of the resources allocated
to its host in $L_1$, which in turn should only receive a percentage $\beta$ of the machine's
total resources, then these two fractions should compound and our guest should end up with
$\alpha\beta$ of the total capacity available.  However, in reality Turtles multiplexes
all guests at all levels onto $L_0$, so in many ways the guest OS and its hypervisor will
be treated equally, rather than as parent and child.  Multi-level device assignment
for Port I/O and Memory-mapped I/O is particularly agnostic to fairness concerns, because
it works exactly the same way as single-level device assignment, and so doesn't care whether
a guest runs at $L_1$ or $L_{100}$ - all guests get equal treatment.  Furthermore, no
experimentation at all is done to analyze the effects of this model on fairness;
the authors even make the statement that this is a secondary concern,
and so they are content with a sort of best-effort.

@@

\subsection{Para-virtualization and Full Virtualization}
\label{sec:relns/para-full}

The second tension seems to arise between the paradigms of \emph{para-virtualization}
and \emph{full virtualization}, which were introduced in Section \ref{sec:background}.
Corporately-funded literature seems to have an eye more toward using as much
commodity software as possible here, for the sake of lower adoption costs, while
academic papers seem more willing to consider modifying client software in
appropriate situations.

The split here is pretty clear as well: Denali, Xen, and Time all consider
para-virtualization to be beneficial and even necessary in the right situations.
In Denali, the isolation kernel presents a modified and simplified hardware abstraction
to its clients, which cleverly allows for some big wins in efficiency and performance
as a side effect.  Xen claims that para-virtualization is necessary to ensure fairness
and prevent starvation in process and I/O scheduling; one could argue, however, that
this is no longer a concern due to the advent of newer virtualization-aware hardware
(see Section \ref{sec:relns/hard-emul}).  The Time paper goes a step further and claims that
even with the new hardware, making the guest OS aware of the virtualized context it
runs in is necessary to ensure the correctness of certain time-dependent applications,
such as systems measurement, high-speed trading, and even video gaming.

In contrast, the Turtles project argues for full virtualization and unmodified guest
OSes and hypervisors.  In the paper, the argument for this goes back to the
motivating goal of performance, though it was unclear to us why para-virtualization
would have any negative impact here.  But it seems also that for any sort of large-scale
commercial deployment of server virtualization technology, the ability to use commodity
guests would be crucial in order to make adoption easy, and we consider this to also
be a likely factor influencing the design choice.

The mClock paper doesn't really fit into these two classifications, since it mostly
deals with the interactions between the hypervisor and its devices, rather than the hypervisor
and its guests.  However, the paper does mention that one of the reasons this disk
scheduling problem is so difficult is because guest OSes have their own algorithms
for efficient disk scheduling, which generally don't play nice with each other.  If it
were possible to modify the guest OS to disable these algorithms, so that the hypervisor
could handle all disk scheduling decision, the difficulty of this problem could likely
be reduced.  This seems to be an argument in favor of para-virtualization.

@@

The final tension we wish to consider is the question of scalability and designing
for the Cloud.  We have observed
that there is a commercial interest in developing super-scalable server virtualization,
largely because it has found such a unique niche with the market growth of Cloud computing
and Infrastructure-as-a-service (IaaS).  While academic research is not uninterested in the question
of scale or the Cloud use case, we do not feel that they are as much of a driving factor in this realm.

We see this trend in the papers we have been discussing as well.  The Turtles paper presents
the motivation for nested virtualization using the argument of its suitability for
Infrastructure-as-a-service, as the two are essentially made for each other.  The mClock paper,
although it does not particularly market itself
for scalability, is motivated by the Cloud-enabled scenario of many virtualized clients
needing to schedule the use of a shared network disk.  Furthermore, the guarantees mClock
makes about fairness, reservations, and proportional share are especially important
in this scenario, as they allow IaaS providers to keep service-level agreements with
their clients.  We find this particularly interesting, as it seems to indicate
that in the Cloud, fairness and performance may find themselves on equal footing in
terms of importance, despite what we have observed in Section \ref{sec:relns/perf-fair}.

Interestingly, the Denali paper also has a particular eye to scalability, and a large
portion of the evaluation section is devoted to demonstrating that the isolation kernel
performs well even when supporting over 10,000 VMs.  However, even in Denali, scalability
is just one goal amongst many, and the paper does not really consider applications
to Cloud computing, being written several years before the Cloud really became prominent.

@@

Having examined the recent literature along these four axes of tension, it is our belief
that a clear trend has emerged in virtualization research.  We believe that as Cloud computing and
IaaS continue to take off, research in this area will continue to trend toward what
is both popular and financially successful.  We expect that the majority of virtualization
research will continue in the direction of full virtualization over para-virtualization,
and that the Cloud-computing/IaaS scenario will become the driving use case.
However, we think that as a result, fairness and performance will no longer be
competing ends, but will in fact become equal goals in virtualized system design.
This will be necessary in order for Cloud providers to meet their SLAs to their
consumers, and we believe fair-scheduling and resource-isolation research will move
from the shadows to the forefront of both corporate and academic research.
